Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-05 SCP2008/paper/metamonads.tex
a few remarks and beautifications

The rest of this paper is structured as follows: The next section
gives an overview of the design space of LU decomposition algorithms.
Section~\ref{CPS}
introduces code generation in MetaOCaml, the problem of name
generation, and the continuation-passing style (CPS) as a general
solution.  We also present the key monad and the issues of generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
LU decomposition algorithm family as separate
modules.  We discuss related work in
Section~\ref{related} and outline future work. In our
conclusion (Section~\ref{conclusion}) we comment
on programming with aspects and sum up our guiding methodology.
Appendices give samples of the generated code, available in
full at \cite{metamonadsURL}.

@@

Throughout, we assume that the reader is familiar with the basic LU
decomposition algorithm, which factors a invertible matrix $A$ into a unit
lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
such that $A = LU$.  Pivoting adds a unitary matrix $P$ such
that the factorization is now $A = PLU$.  The case of numeric matrices is well
covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
a $PLU$ decomposition with $L$ remaining unit lower-triangular. However, 
$U$ is no longer upper triangular but rather ``staggered''
in the upper triangle.
\oleg{Section XXX demonstrates one particular GE algorithm in detail,
  for further reference.}

@@

The groupings are not entirely orthogonal (for example, in-place
decomposition is possible only for specific domains), yet are useful
as guidance in creating the modular LU generator discussed in 
Section~\ref{functors}.

@@

In Figure~\ref{ourmonad}, the monad (yielding values of type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and a value, and
yields an answer of type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable for the environment
classifiers \cite{taha-environment} (such as the type variable |'c| 
in the type of |retN| in Figure~\ref{ourmonad}).
With type constructors taking more
and more arguments, it becomes increasingly difficult to read and write
types -- which we will be doing extensively when writing module
signatures in Section~\ref{functors}. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these sorts of problems has been suggested by 
Jacques Garrigue on the Caml mailing list.
%% (cited from
%% \url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
%% no date, no exact ref). 
%
We use a single type parameter |'p| to
represent all parameters of our monad (all parameters but the type of
the monadic value |'v|). The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type, as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

Various aspects of our generator will need to keep a state during
code generation (see section~\ref{sec:det} for an example).

@@

One particular domain instance is |IntegerDomain|. The type annotation
|DOMAIN| in the definition of |IntegerDomain| makes the 
compiler verify that the defined structure is indeed of a type |DOMAIN|. 
\jacques{In fact, it seems that this annotation cannot be put there
because that will make type v abstract, which will result in compilation
errors for the "lifted" domain which needs to have this type be
available.}
The annotation may be omitted (see |ZpMake| below), in which
case the compiler will verify the type when we try to use that structure as a
|DOMAIN|. In any case, the errors such as missing ``methods'' or
methods with incorrect types will be caught statically, 
\emph{before} any code generation takes place. The variant
|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint:
that full division
is not available. While the |DOMAIN| type may have looked daunting to
some, the implementation is quite straightforward.  Other domains such
as |float| and arbitrary precision exact rational numbers |Num.num|
are equally simple.

@@

\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
\jacques{I don't think I really want to delete the first half of this
subsection.  I guess we can move that into section 3?  We need to be
careful that section 3 does not become 1/2 the paper!  We might even
consider splitting section 3 into 2 sections?}
We will be using MetaOCaml operations to build code expressions, values
of type |('a,'v) code|. However, we wish to make our generators
abstract over particular code representations. Therefore, when writing
the LU generator (Section~\ref{sec:det} and below), instead of
MetaOCaml brackets and escapes we will be using the code combinators
described below. These combinators are functions that produce and
combine values of the type of \emph{abstract code}.

@@

The type of our containers include the lifted domain |Dom| as one of
the components. This is quite convenient since operations on containers
are usually accompanied by the operations on the retrieved and stored
values. Our containers are parametric over a |DOMAINL|, i.e., functors
from a |DOMAINL| module to the actual implementation of a
container. For example, the following functor defines a matrix
container represented as arrays of rows.

\begin{code}
module GenericArrayContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x n m = .< (.~x).(.~n).(.~m) >.
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  \dots
\end{code}
%
The accompanying code\cite{metamonadsURL} includes containers whose
elements are stored in a 1D array, in a
row-wise (C-like) or column-wise (Fortran-like) modes.

@@

The module of signature |INTERNAL_FEATURES| bundles information
tracking aspects. The latter are not directly selectable by the
user. Rather, they are \emph{functions} of other user choices. The
implementations of the tracking aspects such as |Rank|, |NoRank|,
|PackedLower| are quite similar (and simpler) than the implementation
of the |AbstractDet| and |NoDet| aspects in Section~\ref{sec:det}. The
choice of a particular tracking aspect may depend on all other choices
(e.g., it is not possible to extract the |L| factor if the domain is
not a field). Currently we use preflight checks to ensure consistency
of tracking aspects. Previously~\cite{CaretteKiselyov05} we tried to
implement all our preflight tests at the (module) type level, using
sharing constraints and module computations. That lead to obscure
code, long impenetrable type error messages and made compilation very
slow.  Furthermore, at least in OCaml, type-level computations are not
powerful enough to help us with tasks such as verifying that an integer
is prime, as required in section~\ref{sec:domains}.

@@

It would be interesting to implement a camlp4 extension that 
automates the lifting of (simple) modules as done in 
section~\ref{sec:lifteddomains},
first to the code level, and then to monadic values.  Even more
interesting would be a (typed) extension to MetaOCaml that would 
allow us to write such code with the same guarantees that the rest of
MetaOCaml already affords us.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. The implementation of the monadic notation, |perform|,
was joint work with Lydia van Dijk.

>>>>

The rest of this paper is structured as follows: The next section
gives an overview of the design space of LU decomposition algorithms.
\S\ref{CPS}
introduces code generation in MetaOCaml, the problem of name
generation, and the continuation-passing style (CPS) as a general
solution.  We also present the key monad and the issues of generating
control statements. For the sake of reference, in \S\ref{GE-algorithm} 
we present a particular Gaussian Elimination algorithm,
a hand-written implementation of the standard textbook pseudo-code.
\S\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
LU decomposition algorithm family as separate
modules.  We discuss related work in
\S\ref{related} and outline future work. In our
conclusion (\S\ref{conclusion}) we comment
on programming with aspects and sum up our guiding methodology.
Appendices give samples of the generated code, available in
full at \cite{metamonadsURL}.

@@

Throughout, we assume that the reader is familiar with the basic LU
decomposition algorithm, which factors a invertible matrix $A$ into a unit
lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
such that $A = LU$.  Pivoting adds a unitary matrix $P$ such
that the factorization is now $A = PLU$.  The case of numeric matrices is well
covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
a $PLU$ decomposition with $L$ remaining unit lower-triangular. However, 
$U$ is no longer upper triangular but rather ``staggered''
in the upper triangle.
\oleg{\S\ref{GE-algorithm} demonstrates one particular GE algorithm in detail,
  for further reference.}

@@

The groupings are not entirely orthogonal (for example, in-place
decomposition is possible only for specific domains), yet are useful
as guidance in creating the modular LU generator discussed in 
\S\ref{functors}.

@@

In Figure~\ref{ourmonad}, the monad (yielding values of type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and a value, and
yields an answer of type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable for the environment
classifiers \cite{taha-environment} (such as the type variable |'c| 
in the type of |retN| in Figure~\ref{ourmonad}).
With type constructors taking more
and more arguments, it becomes increasingly difficult to read and write
types -- which we will be doing extensively when writing module
signatures in \S\ref{functors}. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these sorts of problems has been suggested by 
Jacques Garrigue on the Caml mailing list.
%% (cited from
%% \url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
%% no date, no exact ref). 
%
We use a single type parameter |'p| to
represent all parameters of our monad (all parameters but the type of
the monadic value |'v|). The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type, as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

Various aspects of our generator will need to keep a state during
code generation (see \S\ref{sec:det} for an example).

@@

One particular domain instance is |IntegerDomain|. The type annotation
|DOMAIN| in the definition of |IntegerDomain| makes the 
compiler verify that the defined structure is indeed of a type |DOMAIN|. 
\jacques{In fact, it seems that this annotation cannot be put there
because that will make type v abstract, which will result in compilation
errors for the "lifted" domain which needs to have this type be
available.}
\oleg{well, we can say IntegerDomain : DOMAIN with type v = int}
The annotation may be omitted (see |ZpMake| below), in which
case the compiler will verify the type when we try to use that structure as a
|DOMAIN|. In any case, the errors such as missing ``methods'' or
methods with incorrect types will be caught statically, 
\emph{before} any code generation takes place. The variant
|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint:
that full division
is not available. While the |DOMAIN| type may have looked daunting to
some, the implementation is quite straightforward.  Other domains such
as |float| and arbitrary precision exact rational numbers |Num.num|
are equally simple.

@@

\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
\jacques{I don't think I really want to delete the first half of this
subsection.  I guess we can move that into section 3?  We need to be
careful that section 3 does not become 1/2 the paper!  We might even
consider splitting section 3 into 2 sections?}
We will be using MetaOCaml operations to build code expressions, values
of type |('a,'v) code|. However, we wish to make our generators
abstract over particular code representations. Therefore, when writing
the LU generator (\S\ref{sec:det} and below), instead of
MetaOCaml brackets and escapes we will be using the code combinators
described below. These combinators are functions that produce and
combine values of the type of \emph{abstract code}.

@@

The type of our containers include the lifted domain |Dom| as one of
the components. This is quite convenient since operations on containers
are usually accompanied by the operations on the retrieved and stored
values. Our containers are parametric over a |DOMAINL|, i.e., functors
from a |DOMAINL| module to the actual implementation of a
container. For example, the following functor defines a matrix
container as a single array, with elements stored in the
row-major order -- the container used in Fig.\ref{fig:GE-algorithm}.
\begin{code}
module GenericVectorContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v container2dfromvector
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x i j = .< ((.~x).arr).(.~i* (.~x).m + .~j) >.
  let dim2 x = .< (.~x).n >.      (* number of rows *)
  let dim1 x = .< (.~x).m >.      (* number of cols *)
  let mapper g a = match g with
  \dots
\end{code}
%
The accompanying code\cite{metamonadsURL} includes containers whose
elements are stored in a 1D array in a column-wise (Fortran-like)
modes, and container for a matrix represented as arrays of rows.

@@

The module of signature |INTERNAL_FEATURES| bundles information
tracking aspects. The latter are not directly selectable by the
user. Rather, they are \emph{functions} of other user choices. The
implementations of the tracking aspects such as |Rank|, |NoRank|,
|PackedLower| are quite similar (and simpler) than the implementation
of the |AbstractDet| and |NoDet| aspects in \S\ref{sec:det}. The
choice of a particular tracking aspect may depend on all other choices
(e.g., it is not possible to extract the |L| factor if the domain is
not a field). Currently we use preflight checks to ensure consistency
of tracking aspects. Previously~\cite{CaretteKiselyov05} we tried to
implement all our preflight tests at the (module) type level, using
sharing constraints and module computations. That lead to obscure
code, long impenetrable type error messages and made compilation very
slow.  Furthermore, at least in OCaml, type-level computations are not
powerful enough to help us with tasks such as verifying that an integer
is prime, as required in \S\ref{sec:domains}.

@@

It would be interesting to implement a camlp4 extension that 
automates the lifting of (simple) modules as done in 
\S\ref{sec:lifteddomains},
first to the code level, and then to monadic values.  Even more
interesting would be a (typed) extension to MetaOCaml that would 
allow us to write such code with the same guarantees that the rest of
MetaOCaml already affords us.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. The implementation of the monadic notation, |perform|,
was joint work with Lydia van Dijk.
Many helpful suggestions by the anonymous reviewers are gratefully
acknowledged.
